### 2019-12-19 manuscript/manuscript.rmd
update manuscript

^3^Department of Biological Sceince, Vanderbilt University
\* Corresponding author, casey.dunn@yale.edu

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, matrix composition and overlappin and outgroup choice) that potentially cause the incongruent results from different studies by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices from differnet studies. Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between site-homogenious and site-heterogeneous models. Our results provide a comprehensive overview of current understanding of the anima-root position and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither. As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.

The exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include:

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model.

To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogeneous models are used.

@@

High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]). Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes). Surprisingly, gene composition and matrice overlap across different studies has been rarely assessed, since most studies relied on their own bioinformatic pipeline for determination of OGs.

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered only when site-heterogeneous CAT models were used. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the animal root position. Feuda et al. 2017 [@feuda2017improved] has proposed to recode the full set of twenty amino acids into six groups that have more frequent evolutionary changes within groups than between groups based on different exchangeability matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they found that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies [@hernandez2019six]. Thus, it is important to evaluate the performance of data recoding method verses non-recoding approaches in past empritical matrices .

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, some earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position, and it should be noted that Whelan2017 [@whelan2017ctenophore] and Simion2017 [@simion2017large] have much broader taxon sampling than other data matrices: Simion2017 has the most taxon sampling, and Whelan2017 contains greater ctenophore taxon sampling.

@@

Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes to tackle the incongruence of the topology among individual gene trees. The basic idea is that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination]. However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring]. Thus, it is important to evaluate the gene composition and matrix overlap to explore congruence of gene sampling across different matrices. It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones.

We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies. Morever, the BUSCO dataset also are frequently used as phylogenomic markers in a varietly of studies. Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conserveness and high-expression properties. However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

We furthered BLAST gene partitions to SwissProt database

Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrice with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). These results indicated that different studies essentially sampled different regions of the genomes with little overlap. One thing should be noted that

@@

Here we summarized all the 167 previous phylogenetic analyses from all studies (Fig. support_published_analyses and Table perviouly_published_analyses). We found that all past phylogenomic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference, recoding methods are used. We found that support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT models. It should be noted that the convergence were not reached or posterior probability is not well supported in many BI analyses. Moreover, the strong support of Porifera-sister are mainly from analyses using Poisson+CAT and recoding methods with site-heterogeneous models. Moreover, only a few past analyses with the most complex GTR+CAT model are converged. However, for the ones that are converged with GTR+CAT model, Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister when choanozoa are used as outgroups.

The use of site-heterogeneous CAT is an infinite mixture model has long been suggested as the best approach to account for systematic errors, particular for data saturation, especially in the case of long-branch attraction (LBA) (Lartillot and Philippe, 2004). The GTR-CAT model often received the best-model fit from cross validation or posterior predictive analyses compared to site-homogeneous or Poisson-CAT models. However, the GTR+CAT model is extremely computational demanding and not feasible in large matrices so many studies only include Poisson+CAT models in their analyses. In contrast, a recent study has suggested that Poisson+CAT model did not outperform site-homogeneous with data partitioning or GTR + CAT models and in some cases can infer incorrect topology based on the simulation analyses. The author also raised the issue of the potential overparameterization in CAT models in some cases. Nevertheless，the debate over the site-heterogeneous vs. site-homogeneous models and/or inclusion or exclusion more distantly related outgroups, constitutes the probably one of the most controversial nodes in the animal tree of life.

@@

We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree. We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.

With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister, every analysis conducted herein choice strongly supported the ctenophore-sister hypothesis, even with site-heterogeneous C models with the most closely related outgroups.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. iqtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT models in every matrix with only Choanoflagellata as outgroup. It should be noted that most Phylobayes runs were converged, several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, and Whelan2015. Moreover, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model. Consistently with past study, we found strong support of Ctenophora-sister in all anaylzed Opisthokonta matrices, indicating that Ctenophora-sister is also favored in all matrices with more distantly related taxa are used as outgroups.

@@

To further explore the effect of site-homogeneous or site-heterogeneous models and outgroup choices in on the animal-root position, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with two outgroup choice (Choanimalia and Opisthokonta), including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis. These results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis.

Surprisingly, we found significant differences of phylogenetic distribution between site-homogenious and site-heterogeneous models. We found that Ctenophora-sister had the higher proportions of supporting genes when site-homogeneous models were used (52.34% to 66.51%), and the outgroup choice has little impact on the distribution of the support phylogenetic signal. This finding is consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models [@shen2017contentious].

@@

@@

# Supplemental Information

>>>>

^3^Department of Biological Sceince, Vanderbilt University

\* Corresponding author, casey.dunn@yale.edu

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between site-homogenious and site-heterogeneous models. Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the anima-root position and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither. As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary text.
Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as Site homogeneous rates, Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models) . (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model.To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogeneous models are used.

@@

High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]) to explore the root of metazoan phylogeny. Although the number of sampled genes might be different across studies, it should expect that the sampled genes should be conserved across most animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes). Surprisingly, gene composition and matrix overlap across different studies has been rarely assessed, and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text).

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [e.g. @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when site-heterogeneous CAT models were used [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving]. Feuda *et al*. (2017) [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in data matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, everal earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices.

@@

Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogney. The basic idea is that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination]. However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring]. Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices. It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones.

We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conservation and ubiquitous. However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrice with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). These results indicated that different studies essentially sampled different regions of the genomes with little overlap.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. support_published_analyses and Table perviouly_published_analyses). Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table PAPERS). Moreover, three of them are reanalyses papers without generating any new data matrices. We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mianly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. support_published_analyses).

@@

Here we reanalyzed `r nrow(analyses_datasets)` from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets). We also progressively trimmed each data matrix with different level of outgroup choice if applicable. We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except for large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices). We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT models in every matrix with only Choanoflagellata as outgroup. It should be noted that most Phylobayes runs were converged, several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Whelan2015_D20. In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model. We found strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups. Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.

@@

To further explore the effect of model and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices that have incongruent results with two outgroup choice, including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal) and four different models in both ML and BI framework. By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis when using LG+GAMMA and C models and the results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis. However, we found significant differences of phylogenetic distribution between LG and C models. We found that Ctenophora-sister had the higher proportions of supporting genes when site-homogeneous models were used (52.34% to 66.51%), and the outgroup choice has little impact on the distribution of the support phylogenetic signal. This finding is consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models [@shen2017contentious].

@@

### Bias of data-recoding

@@

# Supplemental Information
## Models of molecular evolution

The exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR. While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled. There are a couple approaches that have been used in the studies considered here:
